---
title: "Effective Sample Size Under ANCOVA"
author: "Hongtao Zhang"
date: "2023-02-08"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This write-up explores the performance of expected local-information-ratio (ELIR) effective sample size (ESS) under an ANVOCA model. 

Set up the stage. JAGS is required for Bayesian computing. 
```{r, message=FALSE, warning=FALSE}
set.seed(20230208)
library(RBesT)
library(rjags)
library(mvtnorm)
```

Consider a two arm pediatric trial in which the endpoint is the change from baseline at a certain time point. A negative value indicates a treatment benefit. We simulate both the baseline and post-baseline values from a bivariate normal distribution respectively in the treatment and control group. 
```{r}
#total sample size
n_total <- 120
#randomization ratio: treatment:control
RRatio <- 2 
#baseline and post-baseline mean
ped_means_pbo <- c(8.0, 8.0) 
ped_means_active <- c(8.0, 8.0)
#covariance matrix, same for treatment and control arms
ped_vcov <- matrix(c(1.0^2, 0.4, 0.4, 1.43^2), nrow=2)
```

Simulate the data and calculate the change from baseline. 
```{r}
pbo_dat <- rmvnorm((1/(1+RRatio))*n_total, ped_means_pbo, ped_vcov)
active_dat <- rmvnorm((RRatio/(1+RRatio))*n_total, ped_means_active, ped_vcov)
total_dat <- rbind(cbind(pbo_dat, rep(0, (1/(1+RRatio))*n_total)), cbind(active_dat, rep(1, (RRatio/(1+RRatio))*n_total)))

alldt <- as.data.frame(cbind(total_dat, total_dat[,2]-total_dat[,1]))
colnames(alldt) <- c('Base', "Aval", 'TRTP', 'Chg')

head(alldt)
```

The change from baseline is assumed to follow a normal distribution with fixed standard deviation $\sigma$, calculated from the data. A Bayesian ANVOCA model is fitted 
$$E(Chg) = \beta_0 + \beta_1*Base + \delta*Trt.$$
The BUGS code for this model: 
```{r}
jags.script <- "
  model{
    for(i in 1:n) {
      y[i] ~ dnorm(mu[i], 1/ped_var)
      mu[i] <- beta0 + beta1*x[i,1] + delta_p*x[i,3]
    }
    beta0 ~ dnorm(0, 1/10000)
    beta1 ~ dnorm(0, 1/10000)
    delta_p ~ dnorm(adult_mean, 1/adult_var)
  }
"
```

Non-informative nomral priors are used for $\beta_0$ and $\beta_1$. The parameter of interest is the treatment effect $\delta$, for which a normal prior is derived from adult data. 
```{r}
prior_m <- -0.6
prior_s <- 0.6
(prior_mix <- mixnorm(c(1, prior_m, prior_s), param="ms"))
```

Before the prior ESS can be computed, a "reference scale" must be assgined to the prior distribution *`prior_mix`* as required by *RBesT* package. This quantity is essentially the standard deviation of the underlying endpoint which the prior distribution is for. Since $\delta$ is the difference (treatment versus control) in changes from baseline and we assumed each component has a fixed standard deviation $\sigma$, the reference scale is then $\sqrt{2}\sigma$. With that, the ELIR ESS can be computed. 
```{r, message=FALSE}
sigma(prior_mix) <- sqrt(2)*sd(alldt$Chg)
(prior_ess <- ess(prior_mix, "elir"))
```

We then fit the ANCOVA model, obtain a posterior sample of $\delta$ and derive the posterior ELIR ESS. Note that JAGS use the precision parameterization for normal distributions, as opposed to standard deviation. 
```{r, message=FALSE}
jags_data <- list(x = alldt[,1:3], y = alldt[,4], 
                  n = nrow(alldt),
                  adult_mean = prior_m,
                  adult_var = prior_s^2,
                  ped_var = var(alldt$Chg)
)

jag_obj <- jags.model(textConnection(jags.script),
                      data = jags_data,
                      n.chains = 3, 
                      n.adapt = 2000,
                      quiet = T
)

jags_sample <- coda.samples(model=jag_obj,
                            variable.names = c('delta_p'),
                            n.iter=12000, 
                            thin = 2
)
tt<- unlist(jags_sample)
```

The posterior distribution of $\delta$ might be unknown. Therefore, It has to be approximated by a weighted mixture of normal distribution before an ESS can be computed. We also need to assign the same reference scale to the mixture. 
```{r, message=FALSE}
(post_hat <- automixfit(tt, type="norm"))
sigma(post_hat) <- sqrt(2)*sd(alldt$Chg)
(post_ess <- ess(post_hat, "elir"))
```

**The difference between posterior and prior ESS does not equal the sample size `r n_total`. In simulation (500 iterations), the mean difference is 53. ELIR ESS behaves very similar to other ESS definitions, namely, Morita and moment. The predictive consistency property does not seem to hold. **
```{r}
post_ess - prior_ess
```

